The genomic impacts of drift and selection for hybrid performance in maize

### 2015-08-26 main.tex
updated genetic position numbers in the text.

\subsection*{Haplotype phasing}
Although the genotypes of the plants from each population are unphased, the homozygous genotypes of the founders and derived inbreds provide excellent prior information for a probabilistic estimation of genotype phase in the populations. 
We therefore used fastPHASE \citep{scheet2006fast} to estimate the genotype phase of each plant. To estimate the error in phasing, we created test cases by combining the genotypes of two derived inbreds into a hypothetical “F1 hybrid” of unknown phase. 
This F1 was presented to fastPHASE with the rest of the data, except that its parent inbreds were removed. 
Analyses of several hypothetical F1’s from different cycles of selection revealed very low phasing error rates (Table S3). 
Therefore the phased genotypes of cycle 0 plants were used as the starting data for simulations (see below).

@@

No new genetic material was intentionally introduced into either population after the experiment's inception, so the substantial increase in genetic distance could only arise from the loss of genetic diversity within each population. 
Consistent with previous studies of the Iowa RRS \citep{messmer1991genetic, labate1997molecular,  hagdorn2003molecular, hinze2005population}, genome-wide genetic diversity (expected heterozygosity, H) decreases steadily across cycles of selection in both populations (Figure \ref{fig:decline}). \jri{do all these citations make sense here? I just copied these from the abstract}
The loss of heterozygosity is smaller when the two populations are considered together, indicating the loss of different alleles within BSCB1 and BSSS. 
This genetic differentiation is reflected by the tenfold increase in $F_{ST}$ between the founder lines and the populations at cycle 16.

@@

We noticed an irregular increase in the number of polymorphic markers between BSSS cycles 4 and 8 (Table \ref{tab:s2}). 
All of these newly polymorphic markers were present at extremely low frequency and were spread among various individuals. 
This may represent a series of minor alleles that were not captured in our sample of cycle 4 individuals and thus appeared to ‘resurface’ at cycle 8. 
Alternatively, the pattern may be the result of minor contamination at some point in the population’s history. 
It was observed that an allele of the sugary gene associated with sweet corn appeared in the population at this time (O.S. Smith, personal communication), suggesting contamination may be the cause. 
However, the low frequency of the ‘new’ alleles (typically only one or two alleles out of 72 possible in 36 diploid samples) means their effect on population diversity is minimal.  
We did not attempt to incorporate this contamination into our simulation approaches, as it only makes our tests for low heterozygosity slightly more conservative.

\subsection*{Fixation of large genomic regions}
Figure \ref{fig:heterotic} shows heterozygosity varying along the genome at cycle 16 of each population. 
Of particular note are extremely large pericentromeric regions of zero or near-zero heterozygosity spanning tens of megabases. 
These regions experience low rates of meiotic recombination, which creates an expanded physical map relative to their genetic length \citep{ganal2011a-large}. 
In general, the majority of fixed haplotype segments are small ($<2$ cM) in genetic space regardless of their physical size; one exception is 7 cM region on chromosome 1 in the BSCB1 population.

@@

The sheer physical size of the pericentromeric regions yields extremely high marker density on the genetic map, allowing for clear resolution of haplotype phasing and recombination breakpoints. 
To further examine the fixation in these regions, we computationally imputed haplotype phase in the BSSS and BSCB1 populations and used the phased data to track haplotype frequencies and founder of origin. 
In most cases, these fixed haplotypes can be traced back to single founder inbreds. 
For example, in BSSS, a 60 Mb (1.2 cM) region of chromosome 9 became fixed by cycle 12 and traces back to the founder Os420, and in BSCB1, a 60 Mb (0.7 cM) region became fixed on chromosome 4 and traces back to A340 (Figure \ref{fig:genphys}).
Table \ref{tab:fix} gives a summary of the large genomic regions that have become fixed or nearly fixed by cycle 16. 
These regions represent blocks of linked loci that show no evidence of recombination since at least the development of the founding inbred lines in the 1920’s and 1930’s.

>>>>

\subsection*{Haplotype phasing}
Although the genotypes of the plants from each population are unphased, the homozygous genotypes of the founders and derived inbreds provide excellent prior information for a probabilistic estimation of genotype phase in the populations. 
We therefore used fastPHASE \citep{scheet2006fast} to estimate the genotype phase of each plant. To estimate the error in phasing, we created test cases by combining the genotypes of two derived inbreds into a hypothetical “F1 hybrid” of unknown phase. 
This F1 was presented to fastPHASE with the rest of the data, except that its parent inbreds were removed. 
Analyses of several hypothetical F1’s from different cycles of selection revealed very low phasing error rates (Table S3).  \jpg{S2 now?}
Therefore the phased genotypes of cycle 0 plants were used as the starting data for simulations (see below).

@@

No new genetic material was intentionally introduced into either population after the experiment's inception, so the substantial increase in genetic distance could only arise from the loss of genetic diversity within each population. 
Consistent with previous studies of the Iowa RRS \citep{messmer1991genetic, labate1997molecular,  hagdorn2003molecular, hinze2005population}, genome-wide genetic diversity (expected heterozygosity, H) decreases steadily across cycles of selection in both populations (Figure \ref{fig:decline}). %\jri{do all these citations make sense here? I just copied these from the abstract}
The loss of heterozygosity is smaller when the two populations are considered together, indicating the loss of different alleles within BSCB1 and BSSS. 
This genetic differentiation is reflected by the tenfold increase in $F_{ST}$ between the founder lines and the populations at cycle 16.

@@

We noticed an irregular increase in the number of polymorphic markers between BSSS cycles 4 and 8 (Table \ref{tab:s2}). \jpg{S3 now?}
All of these newly polymorphic markers were present at extremely low frequency and were spread among various individuals. 
This may represent a series of minor alleles that were not captured in our sample of cycle 4 individuals and thus appeared to ‘resurface’ at cycle 8. 
Alternatively, the pattern may be the result of minor contamination at some point in the population’s history. 
It was observed that an allele of the sugary gene associated with sweet corn appeared in the population at this time (O.S. Smith, personal communication), suggesting contamination may be the cause. 
However, the low frequency of the ‘new’ alleles (typically only one or two alleles out of 72 possible in 36 diploid samples) means their effect on population diversity is minimal.  
We did not attempt to incorporate this contamination into our simulation approaches, as it only makes our tests for low heterozygosity slightly more conservative.

\subsection*{Fixation of large genomic regions}
Figure \ref{fig:heterotic} shows heterozygosity varying along the genome at cycle 16 of each population. 
Of particular note are extremely large pericentromeric regions of zero or near-zero heterozygosity spanning tens of megabases. 
These regions experience low rates of meiotic recombination, which creates an expanded physical map relative to their genetic length \citep{ganal2011a-large}. 
In general, the majority of fixed haplotype segments are small ($<2$ cM) in genetic space regardless of their physical size; one exception is an 8 cM region on chromosome 1 in the BSCB1 population.

@@

The sheer physical size of the pericentromeric regions yields extremely high marker density on the genetic map, allowing for clear resolution of haplotype phasing and recombination breakpoints. 
To further examine the fixation in these regions, we computationally imputed haplotype phase in the BSSS and BSCB1 populations and used the phased data to track haplotype frequencies and founder of origin. 
In most cases, these fixed haplotypes can be traced back to single founder inbreds. 
For example, in BSSS, a 60 Mb (~2 cM) region of chromosome 9 became fixed by cycle 12 and traces back to the founder Os420, and in BSCB1, a 60 Mb (~3 cM) region became fixed on chromosome 4 and traces back to A340 (Figure \ref{fig:genphys}).
Table \ref{tab:fix} gives a summary of the large genomic regions that have become fixed or nearly fixed by cycle 16. 
These regions represent blocks of linked loci that show no evidence of recombination since at least the development of the founding inbred lines in the 1920’s and 1930’s.
